Predicting Spotify Audio Features from Last.fm Tags

### 2023-06-03 paper/main.tex
improving sections 2 3 and 4

@@

In the remainder of the article, we explain the data gathering and preparation process, as well as the data input formats and varios models.
We will explore various models for the same track and provide insights on how accurately the prediction can be, by using only Last.fm tags.

@@

@@

Several studies have used Last.fm to predict music sentiment, mood, and even audio features.
For example, Laurier et al. analyze how Last.fm tags categorize mood.
In their study, they created a semantic mood space based on Last.fm tags \cite{laurier2009music}.

@@

{\c{C}}ano and Morisio discuss how they create a dataset of music
lyrics annotated with Last.fm tags.
In the creation process, they conclude that Last.fm tags are mostly related to music genre
and positive moods \cite{ccano2017music}.

In a similar direction, Bod{\'o} and Szil{\'a}gyi generate a dataset for lyrics genre classification
by combining the Last.fm with \emph{MusicBrainz} data \cite{bodo2018connecting}.
MusicBrainz is an online database of music editorial metadata\footnote[1]{https://musicbrainz.org/}.

So far, The \emph{Last.fm dataset} has been the most widely used Last.fm dataset\footnotemark[2] in research.
This dataset is a complementary dataset of the Million Song Dataset (MSD) \cite{Bertin-Mahieux2011}.

@@

@@

% Spotify, one of the leaders in the music streaming industry...

The Spotify audio features have been used in multiple studies.
Historically, these features were called the \emph{Echo Nest audio features}.
\emph{The Echo Nest} was an online music intelligence platform that provided users and clients with music analysis services.
Among these services, the Echo Nest offered a database and an API to retrieve audio features for each of the tracks in the database \footnote[3]{https://en.wikipedia.org/wiki/The\_Echo\_Nest}.
Spotify acquired the Echo Nest in 2014.
As a result, the Echo Nest API was shut down and Spotify made these audio features available via the Spotify API.

Nowadays, published research is available under the two terms.
Whereas the most recent studies refer to these audio features as the Spotify audio features,
earlier studies use the Echo Nest denomination.

@@

While Spotify provides a description of the audio features, how they compute or estimate these values is not publicly available.

Wang and Horv{\'a}t use audio features to study differences between male and female artists \cite{wang2019gender}.

Jamdar et al. used EchoNest audio features, combined with lyrics data to classify songs into emotion tags.
These classes were first defined based on a Last.fm tags emotion mapping \cite{jamdar2015emotion}.

@@

Panda and Redinho explore the use of Spotify high-level features applied to Music Emotion Recognition (MER) \cite{panda2021does}.
In particular, they identify that the energy, valence, and acousticness values, provided by the Spotify API,
are highly relevant for emotion classification.
They also achieve better performance on MER models by using their own top-100 features, and they determine that,
although these three Spotify features are relevant in terms of characterizing emotion, more features are needed for MER.

Publicly available datasets of Spotify audio features can be found online,
as a result of open-source and research communities collecting data from the Spotify API and publishing the results.
It is unclear, however, whether these published datasets violate the Spotify API terms of service.

@@

Another example is the \emph{Spotify Audio Features} Kaggle dataset\footnote[4]{https://www.kaggle.com/datasets/tomigelo/spotify-audio-features}.
This dataset contains more than 116,000 unique tracks, and includes audio features for each track.

@@

% JAIME
% I have posted a question on the Spotify Developer forum, relative to the release of a dataset
% https://community.spotify.com/t5/Spotify-for-Developers/May-I-publish-an-open-source-dataset-that-contains-Spotify-Audio/td-p/5529259

Before conducting experiments to predict audio features from tags,
we constructed a dataset, by gathering the data from the Last.fm and Spotify APIs.

@@

This work is scoped within our single-user research line \cite{ramirez2022user}.
In this area, we explore the development of music recommender systems that
characterize the music preferences and listening context only for a single user.
By training our system with single-user data, we also raise the following question: Is it possible to train
recommender systems, and in particular, user-centric systems, by using a single-user dataset?

The user data for this work has been extracted from the
listening history of the corresponding author user in Last.fm\footnote[5]{https://www.last.fm/user/jimmydj2000}.

Similar to other intelligent systems, recommender systems
must be trained, by using user preference data, to produce
suitable recommendations. For this study,
we leverage the knowledge discovery potential
of large historical listening logs, gathered for the mentioned user from Last.fm.

Given the objective of our experiments,
we created a dataset of Last.fm tags and Spotify audio features, indexed by track, by following these steps:

Last.fm is an online music service for users to keep track of their music listening habits.
Last.fm is also an online community where users tag artists, albums, and tracks, according their own taste and perception.

Users apply these tags to
categorize music from their own perspective, which
means that tags do not fit into any structured ontology
or data model.
Tags can refer to any aspect that users consider as a valid descriptor, such as genre,
emotion, or user listenting context.

For nearly two decades, users have been contributing to Last.fm by tagging tracks, albums, and artists with text labels.
Although many of these descriptions are single-worded (e.g \emph{rock}, \emph{dance}, or \emph{happy}),
users can also use short sentences to define a song, such as \emph{I like this track}, or \emph{on the beach}.

\subsubsection{Last.fm Downloaded Data}

Last.fm uses the term \emph{scrobble} to refer to the action of playing a track at a specific moment in time.
Initially, Last.fm monitored user listening activity with a desktop application called \emph{Scrobbler}.
Users used to install this application on their computers to monitor their activity on players such as Winamp or iTunes.
With the advent of music streaming services, the possibilities for users to scrobble their music habits expanded.
Integrations where developed to integrate the scrobbler into popular platforms,
such as Spotify, YouTube, or SoundCloud.
Mobile versions of the Scrobbler were developed for Android and iOS devices, and also open source initiatives flourished\footnote[6]{https://github.com/elamperti/OpenWebScrobbler}.

@@

\begin{itemize}
\item Playback timestamp.
\item Track name.
\item Artist name.
\item Track tags. If the track does not have any tags assigned,
then artist tags were used.
\end{itemize}

Last.fm maps each track (and artist) to a list of community-contributed tags.
For each track-tag mapping, Last.fm includes
a \emph{count} value, which indicates the popularity of the given tag for the track.
Last.fm normalizes this value in the 0-100 range, so the most popular tag for a track can have a
count value of 100.
For example, if \emph{jazz club} is the most popular tag for a track,
then the track might be probably
associated to the following tuple \verb|(jazz club, 100)|.

Users tipically listen to their favorite tracks several times,
so the amount of unique tracks played is much smaller
than the number of track plays. In this case, the amount of
individual tracks listened is about 20,000 and the number of scrobblings is, approximately, 90,000.
Therefore, the user has listened to each song, approximately, 4.5 times on average.

After gathering the listening history and track tags from Last.fm, and identifying the unique
tracks that represent the user music collection, we
collected Spotify audio features for each one of those tracks.

The Spotify audio features are numerical values that represent high-level audio information computed from a specific
track. These values characterize a track, musically speaking,
by measuring relevant musical aspects.
For example, a danceability value of 0.95 means
that a particular song is highly suitable for dancing.

The features provided by the Spotify API are listed in
Table \ref{table:spotify-features}.
The reader can find further details about each feature in the Spotify API documentation\footnote[5]{
      \url{https://developer.spotify.com/documentation/web-api/reference/\#/operations/get-audio-features}
}.

@@

After gathering the data, we identified that the Spotify API had failed to provide audio features for a portion of the tracks.
Similarly, Last.fm also returned no tags for another subset of the tracks.
To prevent problems with missing values, we decided to filter out these tracks from the dataset, and therefore, from our experiments.
After filtering tracks that were missing Last.fm tags or Spotify audio features,
the dataset resulted in 14,009 samples.
Compared to the original 20,000 unique tracks included in the listening history, approximately 6,000 songs were missing either Spotify or Last.fm data.

The mapping between Last.fm and Spotify tracks is performed on an artist-track basis.
The artist and track name extracted from Last.fm are used as parameters of the Spotify Search API.
The Last.fm API provides a unique identifier, the \emph{MusicBrainz} ID.
The Spotify API, however, does not provide this value.

@@

Considering that the data was gathered from a single user,
we explored the data to verify that the distribution of the Spotify audio features
was comparable to larger, and possibly more balanced, Spotify datasets.
In particular, we verified that the distribution of the features,
described in Table \ref{table:audio-features-stats} and Figure \ref{fig:audio-features-distribution},
was comparable to the distribution of the \emph{Spotify Audio Features} Kaggle dataset.
Our dataset presents similar $\mu$ and $\sigma$ values to the Spotify Audio Features Kaggle dataset,
as illustrated in Table \ref{table:audio-features-stats-kaggle}
and Figure \ref{fig:audio-features-distribution-kaggle}.

@@

\begin{table}[h!]
      \begin{center}
      \begin{minipage}{\textwidth}
      \caption{Audio features description}\label{table:audio-features-stats}%
      \begin{tabular}{@{}lll@{}}
      \toprule
      Feature           & $\mu$ & $\sigma$ \\
      \midrule
      Danceability      & 0.60  & 0.19  \\
      Energy            & 0.63  & 0.23  \\
      Acousticness      & 0.22  & 0.30  \\
      Instrumentalness  & 0.51  & 0.38  \\
      Valence           & 0.44  & 0.28  \\
      \botrule
      \end{tabular}
      \end{minipage}
      \end{center}
\end{table}
\begin{table}[h!]
      \begin{center}
      \begin{minipage}{\textwidth}
      \caption{Audio features description of \emph{Spotify Audio Features} Kaggle dataset}\label{table:audio-features-stats-kaggle}%
      \begin{tabular}{@{}lll@{}}
      \toprule
      Feature           & $\mu$ & $\sigma$ \\
      \midrule
      Danceability      & 0.58   & 0.19  \\
      Energy            & 0.57  & 0.26  \\
      Acousticness      & 0.34  & 0.25  \\
      Instrumentalness  & 0.22  & 0.36  \\
      Valence           & 0.44  & 0.26  \\
      \botrule
      \end{tabular}
      \end{minipage}
      \end{center}
\end{table}

\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{images/features-distribution.png}
      \caption{Distribution of audio features in our single-user dataset.}
      \label{fig:audio-features-distribution}
\end{figure}

\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{images/feature-distribution-kaggle.png}
      \caption{Distribution of audio features in the \emph{Spotify Audio Features} Kaggle dataset.}
      \label{fig:audio-features-distribution-kaggle}
\end{figure}

@@

We trained commonly used machine learning models to predict Spotify audio features, given the set of Last.fm tags for a particular track.

>>>>

% TENSES: https://blog.wordvice.com/video-which-verb-tenses-should-i-use-in-a-research-paper/

@@

In the remainder of the article, we explain the data gathering and preparation process, as well as the data input formats and varios models.
We will explore various models for the same track and provide insights on how accurately the prediction can be, by using only Last.fm tags.
\textcolor{red}{UNDER DEVELOPMENT}

\section{Data Sources}

Last.fm is an online music community where users keep track of their music listening habits.
Users apply tags to artists, tracks, and albums to categorize and describe music from their own perspective, which
implies that the Last.fm tags space does not fit into any structured ontology or data model.
A tag can refer to any aspect that users consider to be a valid descriptor, such as genre,
emotion, or user listening context.

For nearly two decades, many music aficionados have collectively contributed their own unique, personal interpretation, opinions and feelings
as tags in Last.fm.
Although many of these tags are single-worded descriptors (e.g \emph{rock}, \emph{dance}, or \emph{happy}),
users also use short sentences to describe music, such as \emph{I like this track}, or \emph{on the beach}.

Tags are available via the Last.fm API.

Spotify, one of the leaders in the music streaming industry, provides information about the tracks in their catalog via the Spotify developers API.
Among the different data entities exposed, the API provides access to the \emph{Audio Features} for each track.

The Spotify audio features are numerical values that represent high-level audio information computed from a specific
track. These values characterize a track, musically speaking,
by measuring musical aspects that, in many of these features, are related to the user perspective or recommendation factors.
For example, a danceability value of 0.95 means
that a particular song is highly suitable for dancing.

The features provided by the Spotify API are listed in
Table \ref{table:spotify-features}.
While Spotify provides a description of the audio features.
How they compute or estimate these values is not publicly available.
The reader can find further details about each feature in the Spotify API documentation\footnote[5]{
      \url{https://developer.spotify.com/documentation/web-api/reference/\#/operations/get-audio-features}
}.

@@

The objective of this study is to predict Spotify audio features from Last.fm tags, so the objective of this section is to
explore existing literature that is related to the problem and review the facts that are important or relevant for our experimentation.

The use of Last.fm tags and Spotify audio features has been common and prolific in studies
that have applied machine learning to resolve MIR challenges.

An important concept to consider is the arousal-valence cuadrants, present in music emotion recognition, 
related to some studies that will be reviewed in the following sections, and also related to the valence and energy
Spotify audio features.

@@

Several studies have used Last.fm to predict music sentiment, mood, and even audio features.
For example, Laurier et al. analyzed how Last.fm tags categorize mood.
In their study, they created a semantic mood space based on Last.fm tags \cite{laurier2009music}.

@@

{\c{C}}ano and Morisio performed an analysis of Last.fm tags to create a dataset of music
lyrics annotated with Last.fm tags.
In the creation process, they concluded that Last.fm tags are mostly related to music genre
and positive moods \cite{ccano2017music}.

In a similar direction, Bod{\'o} and Szil{\'a}gyi generated a dataset for lyrics genre classification
by combining the Last.fm with \emph{MusicBrainz} data \cite{bodo2018connecting}.
MusicBrainz is an online database of music editorial metadata\footnote[1]{https://musicbrainz.org/}.

Although different datasets that contain Last.fm tags have been created, so far,
the \emph{Last.fm dataset}\footnotemark[2] has been the most widely used in research.
This dataset is a complementary dataset of the Million Song Dataset (MSD) \cite{Bertin-Mahieux2011}.

@@

\textcolor{red}{We introduced this section talking about machine learning but only refeferred to studies
that use Last.fm tags in exploration and data set creation}

@@

Historically, the Spotify audio features features were called the \emph{Echo Nest audio features}.
\emph{The Echo Nest} was an online music intelligence platform that provided users and clients with music analysis services.
Among these services, the Echo Nest offered a database and an API to retrieve audio features for each of the tracks in the database \footnote[3]{https://en.wikipedia.org/wiki/The\_Echo\_Nest}.
Spotify acquired the Echo Nest in 2014.
As a result, the Echo Nest API was eventually deactivated and Spotify migrated these audio features to the Spotify API.

Nowadays, the two terms can be found in published research.
Whereas the most recent studies refer to the Spotify audio features,
earlier studies use the Echo Nest denomination.

@@

Similarly to Last.fm, Spotify (or Echo Nest) audio features are commonly present in MIR research.
For example, Wang and Horv{\'a}t used  these audio features to study the differences 
between male and female artists \cite{wang2019gender}.
\textcolor{red}{Y que consiguieron?}

Jamdar et al. used Echo Nest audio features, combined with lyrics data to classify songs into emotion tags.
These classes were first defined based on a Last.fm tags emotion mapping \cite{jamdar2015emotion}.

@@

Panda and Redinho explore the use of Spotify high-level features applied to Music Emotion Recognition (MER) \cite{panda2021does}.
In particular, they identify that the energy, valence, and acousticness values, provided by the Spotify API,
are highly relevant for emotion classification.
They also achieve better performance on MER models by using their own top-100 features, and they determine that,
although these three Spotify features are relevant in terms of characterizing emotion,
more features are needed to effectively recognize emotion.
Another interesting observation of this study is the the identification of energy as a surrogate of arousal valence

Similarly to Last.fm tags, Spotify audio features can be found in a number of datasets.
Publicly available datasets of Spotify audio features can be found online,
as a result of open-source and research communities collecting data from the Spotify API and publishing the results.
It is unclear, however, whether these published datasets completely meet the Spotify API terms of service.

@@

Another publicly available dataset is the \emph{Spotify Audio Features} Kaggle dataset\footnote[4]{https://www.kaggle.com/datasets/tomigelo/spotify-audio-features}.
This dataset contains more than 116,000 unique tracks, and includes audio features for each track.

For this reason, we chose to generate our own dataset.

@@

Before conducting experiments to predict audio features from tags,
we constructed a dataset of Last.fm tags and Spotify audio features, indexed by track, by gathering the data from the Last.fm and Spotify APIs.

The tracks were selected from the listening history of a single user.

@@

This work is scoped within our single-user research line \cite{ramirez2022user}.
In this area, we explore the development of music recommender systems that
characterize the music preferences and listening context only for a single user.
Therefore, we extracted the data from the
listening history of the corresponding author in Last.fm\footnote[5]{https://www.last.fm/user/jimmydj2000}.

By training our system in a single-user space, we also raise the following question: Is it possible to train
recommender systems, and in particular, user-centric systems, by using a single-user dataset?
Additionally, we wanted to explore the idea of mimicking the fact that each human perceives music individually.
If we train a system on data from different users, then the system would share the view of multiple individuals.

Using a single-user data set might sound counterproductive in a machine learning scenario, specially considering
how machine learning breakthroughs have attempted, and succeeded in many cases, to generalize in a particular problem.
This is not objective of this study, which explores how a machine learning model can represent the music consumption experience of a single human.
The model must be able to generalize, but only within the context of the user's musical taste, which be believe can be possible, given a sufficiently large listening history.

\subsection{Gathering Listening History and Tags from Last.fm}

Last.fm uses the term \emph{scrobble} to refer to the action of playing a track at a specific moment in time.
Last.fm started monitoring user listening activity with a desktop application called \emph{Scrobbler}.
Users used to install this application on their computers to monitor their activity on players such as Winamp or iTunes.
With the advent of music streaming services, the possibilities for users to scrobble their music habits expanded.
Integrations where developed to integrate the scrobbler into popular platforms,
such as Spotify, YouTube, or SoundCloud.
Mobile versions of the Scrobbler were also developed for Android and iOS devices, while open source initiatives flourished too\footnote[6]{https://github.com/elamperti/OpenWebScrobbler}.

@@

\begin{itemize}
\item Playback timestamp
\item Track name
\item Artist name
\item Track tags
\end{itemize}

Last.fm maps each track (and artist) to a list of community-contributed tags.
For each track-tag mapping, Last.fm includes
a \emph{count} value, which indicates the popularity of the given tag for the track.
Last.fm normalizes this value in the 0-100 range, so the most popular tag for a track is tipically associated with a
count value of 100.
For example, if \emph{jazz} is the most popular tag for a track,
then the track might be probably associated to the following tuple \verb|(jazz, 100)|.

Users tipically listen to their favorite tracks several times,
so the amount of unique tracks played is smaller
than the number of track plays. In this case, the amount of
individual tracks listened in a 15-year period is about \num{20000} and the number of scrobblings is, approximately, \num{90000}.
Therefore, the user has listened to each song, approximately, \num{4.5} times on average.

\subsection{Gathering Spotify Audio Features}

After collecting the listening history and track tags from Last.fm, and identifying the unique
tracks that represent the user music collection, we used the Spotify API to
collect audio features for each one of the \num{20000} individual tracks.

The mapping between Last.fm and Spotify tracks was performed on an artist-track basis.
For each track, the artist and track name extracted from Last.fm were used as parameters of the Spotify Search API.
The Last.fm API provides a unique identifier, the \emph{MusicBrainz} ID.
The Spotify API, however, does not provide this value.

@@

After retrieving audio features, we identified that the Spotify API had failed to provide audio features for a portion of the tracks.
Similarly, Last.fm also returned no tags for another subset of the tracks.
To prevent problems with missing values, we decided to filter out these tracks from the dataset.
After filtering tracks that were missing Last.fm tags or Spotify audio features,
the dataset resulted in \num{14009} samples.
Compared to the original \num{20000} unique tracks included in the listening history, approximately \num{6000} songs were missing either Spotify or Last.fm data.
In other words, about 70\% of the tracks in the user listening history included relevant information for the study.

@@

Considering that the data was gathered from a single user,
we explored the data to verify that the distribution of the Spotify audio features
was comparable to larger, and possibly more balanced, Spotify datasets.
In particular, we verified that the distribution of the features,
described in Table \ref{table:audio-features-stats} and Figure \ref{fig:audio-features-distribution},
was comparable to the distribution of the Spotify Audio Features Kaggle dataset.
Our dataset, which we call \emph{Jimmydj2000 Single-user} dataset, presents mean ($\mu$) and standard deviation($\sigma$) values that are comparable to the same values of the Spotify Audio Features Kaggle dataset,
as illustrated in Table \ref{table:audio-features-stats-kaggle}
and Figure \ref{fig:audio-features-distribution-kaggle}.

@@

\begin{table}[!ht]
      \begin{minipage}{.5\linewidth}
      \caption{Audio features description of the Jimmydj2000 Single-user dataset.}\label{table:audio-features-stats}
            \centering
                  \begin{tabular}{@{}lll@{}}
                  \toprule
                  Feature           & $\mu$ & $\sigma$ \\
                  \midrule
                  Danceability      & 0.60  & 0.19  \\
                  Energy            & 0.63  & 0.23  \\
                  Acousticness      & 0.22  & 0.30  \\
                  Instrumentalness  & 0.51  & 0.38  \\
                  Valence           & 0.44  & 0.28  \\
                  \botrule
                  \end{tabular}
      \end{minipage}
      \begin{minipage}{.5\linewidth}
      \caption{Audio features description of Spotify Audio Features Kaggle dataset.}\label{table:audio-features-stats-kaggle}%
            \centering
                  \begin{tabular}{@{}lll@{}}
                  \toprule
                  Feature           & $\mu$ & $\sigma$ \\
                  \midrule
                  Danceability      & 0.58   & 0.19  \\
                  Energy            & 0.57  & 0.26  \\
                  Acousticness      & 0.34  & 0.25  \\
                  Instrumentalness  & 0.22  & 0.36  \\
                  Valence           & 0.44  & 0.26  \\
                  \botrule
                  \end{tabular}
      \end{minipage}

\end{table}
% \begin{table}[h!]
%       \begin{center}

%       \end{center}
% \end{table}

\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{images/features-distribution.png}
      \caption{Distribution of audio features in the Jimmydj2000 Single-user dataset.}
      \label{fig:audio-features-distribution}
\end{figure}

\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{images/feature-distribution-kaggle.png}
      \caption{Distribution of audio features in the Spotify Audio Features Kaggle dataset.}
      \label{fig:audio-features-distribution-kaggle}
\end{figure}

@@

We trained three commonly used machine learning models to predict Spotify audio features from Last.fm tags by using the
